---
title: "Tidy data"
subtitle: "structuring datasets to make analysis easy"
author: "Andrew MacDonald"
date: "07 Feb 2017"
output:
  xaringan::moon_reader:
    seal: yes
    nature:
      highlightStyle: github
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# What is "tidy data"

> Happy families are all alike; every unhappy family is unhappy in its own way — Leo Tolstoy

--

1. Each variable forms a column.

--

1. Each observation forms a row.

--

1. Each type of observational unit forms a table.

--

see [the tidy data vignette](ftp://cran.r-project.org/pub/R/web/packages/tidyr/vignettes/tidy-data.html)

---

# an example

```{r comment='#'}
library(gapminder)
head(gapminder)
```

--
</br>
</br>
</br>


**The `data.frame` is a fundamental part of any "tidy" workflow**

---
class: center
# Key principle
</br>
--
<font size = 20> raw data should stay </br> <strong>exactly the way it is</strong> </font>
--
</br>
</br>
this suggests that we will need to rearrange data _automatically_

---

class: center, middle

# Introducing: the most controversial symbol in modern R

---

class: center, middle
</br>
`%>%`
---
# the pipe

```{r eval=FALSE}
foo_foo <- little_bunny()

## orthodox R style
bop_on(scoop_up(hop_through(foo_foo, forest), field_mouse), head)
```
--
```{r eval=FALSE}
## novel pipe syntax
foo_foo %>%
  hop_through(forest) %>%
  scoop_up(field_mouse) %>%
  bop_on(head)
```

---
## dplyr & tidyr

1. `select()`
2. `filter()`
3. `group_by()`
4. `summarize()`
5. `mutate()`

and from `tidyr`
1. `gather`
2. `spread`
3. `separate`
4. `unite`


