---
title: "Introduction to the manipulation of tidy data"
author: "andrew"
date: "February 7, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we load the packages that we need via `library`: 

```{r pkgs}
library(gapminder)
library(dplyr)
library(tidyr)
```

We can use some handy R functions to view this dataset:

```{r eval=FALSE}
head(gapminder)
View(gapminder)
```

> small tip: you can create handy "sections" in your R code within R studio by clicking `Code > Insert section`. That only works when you are in a `.R` file

There are **Five basic functions** for manipulating datasets. These are the five "verbs" of dplyr, and together you can express very sophisticated data transformations with them.

## Reordering data: `arrange`

The simplest operation we can do to a dataset is to change the sequence of the rows. We can do this easily with `arrange` from dplyr.

```{r}
arrange(gapminder, year)
```


## Reducing rows and columns: `select()` and `filter()`

You can select columns based on their names. You can also indicate which columns you do **not** want. You can even select by names.
```{r}
select(gapminder, continent, year, lifeExp)
## remove with -
select(gapminder, -continent)
## remove by range
select(gapminder, lifeExp:gdpPercap)
```

There are other ways of selecting columns. See a complete list at `?select`

Just as `select()` reduces the number of columns, `filter()` reduces the number of rows:

```{r}
filter(gapminder, year > 1983)
## before 1970
filter(gapminder, year < 1970)
filter(gapminder, year > 1970, year < 1990)
```

## Introducing the pipe operator!

`%>%`

Perhaps no recent development in the world of R is as useful, or as controversial, as these four characters. The pipe allows us to "rephrase" the way we write R code. Usually, we write R from the "inside out": 

```r
mean(abs(rnorm(30)))
```

Here, the first function (`rnorm(30)`) is on the inside, while the subsequent functions are outside each set of parentheses. This is difficult to read, because it is not like any human language.

The pipe, however, lets us write this sequentially:

```r
rnorm(30) %>% abs %>% mean
```

### from data unto data

All the dplyr functions we just looked at share one common trait: they all take a `data.frame` as their first argument, and they all return a data.frame as well. This means we can chain them together with the pipe: 

```{r}
gapminder %>% 
  select(continent, country, pop) %>% 
  filter(pop > 1e8)

## adding a new column
gapminder %>% 
  mutate(total_gdp = pop * gdpPercap)

## summarize

## what is the average population for an Asian country over all the time studied

gapminder %>% 
  filter(continent == "Asia") %>% 
  group_by(country) %>% 
  summarize(avg_pop = mean(pop))

## challenge! 
## What is the average total GDP across all continents in 1987

continent_means <- gapminder %>%
  mutate(total_gdp = pop * gdpPercap) %>% 
  filter(year == 1987) %>% 
  group_by(continent) %>% 
  summarise(avg_total_gdp = mean(total_gdp))

continent_means

## tidyr

gapminder %>% 
  select(country, year, pop) %>% 
  spread(key = year, value = pop)

year_by_gdp <- gapminder %>% 
  select(country, year, gdpPercap) %>% 
  spread(year, value = gdpPercap)

year_by_gdp %>% 
  gather(key = new_year, value = new_pop, 2:13)
year_by_gdp %>% 
  gather(key = new_year, value = new_pop, `1952`:`2007`)
year_by_gdp %>% 
  gather(key = new_year, value = new_pop, -country)

## scale pop by subtracting the mean for each country
gapminder %>%
  group_by(country) %>% 
  mutate(scale_pop = pop - mean(pop))
```